A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-22 papers/execution-model/model.tex
final touches

\begin{itemize}
\item An operational semantics for functional programs running
  alongside smart contracts in a blockchain. 
\item Established various properties of the combined system withs
  proofs in upcoming techreport.
\item An implementation of a low-level OCaml-API to the Tezos
  blockchain, which corresponds to the operational semantics.
\end{itemize}

@@

The execution model is an abstraction of the Tezos blockchain \cite{tezos-whitepaper}
augmented with functional (OCaml) programs that interact with the
blockchain. As the blockchain is realized by a peer-to-peer network of
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system. In particular, we do
\textbf{not} formalize the execution of the smart contracts
themselves, we rather consider them as black boxes that execute in a
safe-safe way. Tezos' smart contract language Michelson and its
properties have been formalized elsewhere \cite{type-safe-michelson}.

@@

The local node offers several ways to query the current state of
the blockchain.
\emph{Queries} are defined by the following grammar:
\begin{align*}
  \QRY & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE\ADDR
  \mid \GETSTATUS\OPH
  \mid \GETSTORAGE\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the code of a contract by the
operation hash of its originating transaction.

@@

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR \mid \CAST\EXPR\TYPE\TYPEU
  \\& \mid \QOP\  \EXPR \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  % \\&\mid \CONTRACT\ \TYPE\ \EXPR
  \\& \mid \TOSTRING\ \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= \GETBALANCE{} \mid \GETSTATUS{} \mid \GETSTORAGE{} \mid \GETCODE{} \mid \GETCOUNTER{}
         \mid \GETTYPE{}
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by programs, i.e., closed
expressions of type unit that run on the
local node. Each program can send transactions on behalf of any
account on the local node.  Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types, exceptions and fixpoint. Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which is only applicable to pairs of
types related in the reflexive, transitive, and symmetric closure of
the subtyping relation in Figure~\ref{fig:subtyping}. Upcasts will
just go through, but downcasts may raise an exception, as usual.

@@

We define a relation $\SubType$ that establishes pairs of types that can be
subjected to casting. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.

>>>>

\begin{itemize}
\item An operational semantics for functional programs running
  alongside smart contracts in a blockchain. 
\item Established various properties of the combined system withs
  proofs in upcoming techreport.
\item An implementation of a low-level OCaml-API to the Tezos
  blockchain, which corresponds to the operational semantics.\footnote{%
    Available at \url{https://github.com/tezos-project/Tezos-Ocaml-API}.}
\end{itemize}

@@

The execution model is an abstraction of the Tezos blockchain \cite{tezos-whitepaper}
augmented with functional (OCaml) programs that interact with the
blockchain. As the blockchain is realized by a peer-to-peer network of
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system. In particular, we do
\textbf{not} formalize the execution of the smart contracts
themselves, we rather consider them as black boxes that execute in a
safe-safe way. Tezos' smart contract language Michelson and its
properties have been formalized elsewhere \cite{DBLP:conf/fm/BernardoCHPT19}.

@@

The local node offers several ways to query the current state of
the blockchain.
\emph{Query operators} are defined by the following grammar:
\begin{align*}
  \QOP & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE{}%\ADDR
  \mid \GETSTATUS{}%\OPH
  \mid \GETSTORAGE{}%\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT{}%\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the public hash of a contract by the
operation hash of its originating transaction.

@@

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR \mid \CAST\EXPR\TYPE\TYPEU
  \\& \mid \QOP\  \EXPR \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  % \\&\mid \CONTRACT\ \TYPE\ \EXPR
  \\& \mid \TOSTRING\ \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
            &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR
  % \\
  % \QOP & ::= \GETBALANCE{} \mid \GETSTATUS{} \mid \GETSTORAGE{} \mid \GETCODE{} \mid \GETCOUNTER{}
  %        \mid \GETTYPE{}
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by programs, i.e., closed
expressions of type unit that run on the
local node. Each program can send transactions on behalf of any
account on the local node.  Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types, exceptions and fixpoint. Moreover, there are
domain-specific primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which we describe after discussing types.

@@

Type casts are only applicable to certain pairs of
types governed by a relation $\SubType$, which could also serve as a
subtyping relation. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.
